---
title: "Visualizing the Central Limit Theorem"
author: "Michael Smith"
date: "May 29, 2016"
output:
  html_document:
    keep_md: true
---

First, we need to establish an initial sample and discover it's mean:

```{r}
x <- rnorm(50, 22, 5)
xbar <- mean(x)
summary(x)
```

Next we create a bootstrap sample based of the initial sample as though it was a population:

```{r}
nsims <- 1000
bootnorm <- numeric(nsims)
for (i in 1:nsims) {
	temp <- sample(x, 50, replace=TRUE)
	bootnorm[i] <- mean(temp)
}
summary(bootnorm)
```

Now that we have our sample, we can see on a histogram that the bootstrap sample is normally distributed around the initial sample population mean in red:

```{r}
hist(bootnorm)
abline(v=xbar, col="red")
```

As we increase the number of bootstrap iterations taken, the histogram will become more and more normally distributed. We can also compare it to increased sample sizes of the original population.

```{r}
x <- rnorm(100, 22, 5)
xbar <- mean(x)
summary(x)
nsims <- 5000
bootnorm <- numeric(nsims)
for (i in 1:nsims) {
	temp <- sample(x, 100, replace=TRUE)
	bootnorm[i] <- mean(temp)
}
summary(bootnorm)
hist(bootnorm)
abline(v=xbar, col="red")
```

This effect can also be seen on samples with exponential distributions:

```{r}
x <- rexp(50)
xbar <- mean(x)
summary(x)
nsims <- 1000
bootnorm <- numeric(nsims)
for (i in 1:nsims) {
	temp <- sample(x, 50, replace=TRUE)
	bootnorm[i] <- mean(temp)
}
summary(bootnorm)
hist(bootnorm)
abline(v=xbar, col="red")
```

And of course, larger sample sizes within exponential distributions:

```{r}
x <- rexp(100)
xbar <- mean(x)
summary(x)
nsims <- 5000
bootnorm <- numeric(nsims)
for (i in 1:nsims) {
	temp <- sample(x, 100, replace=TRUE)
	bootnorm[i] <- mean(temp)
}
summary(bootnorm)
hist(bootnorm)
abline(v=xbar, col="red")
```