---
title: "MSmith_9.5_Volitility_402"
author: "Michael Smith"
date: "July 17, 2016"
output:
  html_document:
    keep_md: true
---

##Introduction

In this document, we examine the volatility of the Honeywell International Inc. (HON) stock and compare it to the volatility of the S&P. First I capture the data of the two stocks and establish a function for calculating the volatility.
```{r include=FALSE}
require("tseries")

HONdata <- get.hist.quote('hon',quote="Close")
HONret <- log(lag(HONdata)) - log(HONdata)
HONvol <- sd(HONret) * sqrt(250) * 100

SNPdata <- get.hist.quote('^gspc',quote="Close")
SNPret <- log(lag(SNPdata)) - log(SNPdata)
SNPvol <- sd(SNPret) * sqrt(250) * 100

getVol <- function(d, logrets) {
  var = 0
  lam = 0
  varlist <- c()
  for (r in logrets) {
    lam = lam*(1 - 1/d) + 1
    var = (1 - 1/lam)*var + (1/lam)*r^2
    varlist <- c(varlist, var)
  }
  sqrt(varlist)
}
```

```{r}
volest <- getVol(10,HONret)
volest2 <- getVol(30,HONret)
volest3 <- getVol(100,HONret)

SNPvolest <- getVol(10,SNPret)
SNPvolest2 <- getVol(30,SNPret)
SNPvolest3 <- getVol(100,SNPret)
```

###Volitility Plot Overlay

In this plot we use three different decay values, black is 10, red is 30, and blue is 100.

```{r}
plot(volest,type="l")
lines(volest2,type="l",col="red")
lines(volest3,type="l",col="blue")
```

###Volitility Comparison to SNP

And finally I compare the volitility of HON to S&P at a decay value of 10. HON is Green; S&P is yellow.

```{r}
plot(volest,type="l",col="green")
lines(SNPvolest,type="l",col="yellow")
```